Emittance preservation of an electron bunch in a loaded quasi-linear plasma wakefield

### 2017-07-13 Tex/Paper.tex
End of day commit

\section[\label{S:I}]{Introduction}

% interested from SMI/AWAKE
% requirements for AWAKE Run 2

The preliminary design of AWAKE Run 2 proposes to use two plasma sections. The first section of 4m is the SMI stage. The
electron beam will be injected into the modulated proton beam before stage two, where acceleration will occur. As the
$e_z$ field will decrease due to the gap between the two cells, it is desireable to keep this as short as possible
\cite{adli:2016}.

\section[\label{S:M}]{Method}
%- explain analogy of short-bunch, SMI case, by referring to earlier Veronica-papers
%- simulation setup PIC/quickpic OS

The main focus of this study has been on the beam loading of the electron beam. In order to eliminate other factors that
may affect this, we have tried several approaches to create a stable drive beam structure based on previous SMI studies
[citations].

Our first approach was to use a premodulated, short proton beam with the same structure as a section of the full AWAKE
proton drive beam. These studies were done using the full PIC code Osiris \cite{fonseca:2002} using 2D
cylindrical-symetric simulations. The proton beam was pre-modulated by a clipped cosine function to the longitudinal
density profile, with a period matching the wavelength, $\lambda_p$, of the plasma. The length was limited to
$26\cdot\lambda_p$, and the electron beam injected after the 20th micro-bunch \cite{berglyd_olsen:2015}. We performed
several parameter scans with this setup, testing for optimal charge as well as beam length
\cite{adli:2016, berglyd_olsen:2016}.

[Add something about the optimal results]

In order to evaluate the quality of the beam, we also needed to study its emittance. Full PIC codes like Osiris are
vulnerable to numerical growth of emittance caused by the ``numerical Cherenkov effect'' \cite{godfrey:1974}. This is a
know issue with the Yee EMF solver, which causes the phase velocity of electromagnetic fields to be lowet than $c$,
while the beam moves very close to $c$. The effect can be mitigated somewhat by a the Lehe solver \cite{lehe:2013}, but
the effect is still prominent in the the high density regions of the electron beam.

In order to study the emittance evolution of the beam we used QuickPIC, a fully relativistic 3D PIC code
\cite{huang:2006, an:2013}.

\subsection[\label{S:M:Setup}]{Drive Beam Parameters}

In these simulations we use a single proton drive bunch that sets up a wakefield comparable to that which we expect to
see from the self-modulated SPS beam in AWAKE. The baseline AWAKE drive beam contains $3\nexp{11}$ protons
\cite{gschwendtner:2016}. Only half of these are driving the wakefield, presuming the self-modulation is seeded in the
middle of the unmodulated bunch. In addition. a significant portion of the protons are lost during self-modulation. The
needed gap between the two plasma cells also contributes to a decreased density of protons near the axis, and some of
the field is drained by other protons seeing an accelerating face within the drive beam. In total, the expected peak
$E_{z}$ field of $~2\funit{GV}{m}$ in the first plasma cell drops to $500-600\funit{MV}{m}$ in the second cell
\cite{awake_collaboration:2016}. A single proton beam of $1.46\nexp{10}$, or $2.34\unit{nC}$ and $7\unit{kA}$, is needed
to drive an $E_{z}$ field of a comparable magnitude.

The single bunch approach was chosen to reduce simulation time for the 3D simulations, and to create a stable
environment for the witness beam. We are not considering the evolution of the proton beam in this test setup, so to
prevent the proton beam from evolving radially, we increased its mass.

The basline AWAKE drive beam current is insufficient to produce a plasma bubble. The plasma electrons are depleted to
around $65\%$ of nominal plasma density at the injection point of the electron beam. This condition is replicated in out
single bunch case. The single bunch setup also uses the baseline proton energy $W_{pb} = 400\unit{GeV}$.

\subsection[\label{S:M:Setup}]{Witness Beam Parameters}

The witness beam in our simulation differs from AWAKE baseline parameters on several key points. Initial beam energy is
set such that $\gamma_{eb} = \gamma_{pb} = 426.3$. This is to eliminate the problem of initial de-phasing of the witness
beam. AWAKE baseline witness beam energy for Run 1 is $W_{eb} = 10-20\unit{MeV}$ \cite{gschwendtner:2016}. Beam loading
of a short witness beam is sensitive to initial position. AWAKE Run 2 will likely require a higher initial energy, but a
$W_{eb} = 30--50\unit{MeV}$ is sufficient to minimise de-phasing \cite{berglyd_olsen:2015}. For this case we have
eliminated it entirely in order to reduce the number of variables in our parameter scans.

Earlier simulations with witness beam $\sigma_{r} = 100\unit{\mu m}$ causes the beam radius to rapidly shrink as the
beam enters the plasma \cite{berglyd_olsen:2016}. This is due to a mismatch between beam emittance and the plasma
density. The relation is given by
\begin{align}
    \beta = \frac{\sigma_r^2}{\epsilon_g} = \frac{\lambda_p}{2\pi}\sqrt{2\gamma_{rel}}, \label{EQ:Matched}
\end{align}
where $\lambda_{p}$ is the plasma wavelength. In these simulations we have used matched beams for emittances between
$2$ and $6\unit{\mu m}$.

\section[\label{S:BL}]{Beam Loading}

%~ Eric's Comments
%~ Sec N: Beam loading [in this section we describe the interesting physics, based on a single parameters set]

@@

The structure of the single drive beam setup behaves similarly to the self-modulated case. However, since the drive
beam is prevented from significant evolution in this simulation setup, we are presented with an idealised case where
the electron witness beam sees consistent fields.

\begin{figure}
    \includegraphics[width=0.99\linewidth,trim={4mm 0mm 4mm 0mm},clip]{figures/beamLoading}
    \label{Fig:BeamLoading}
    \caption{Comparison between the unloaded longitudinal e-field (no witness beam) and the loaded e-field along the
        axis. The magnitude of the beam density along the axis is shown for reference. The beams travel towards the
        left. $\xi = z - tc$ is the position in the simulation box. The beam density is in units of plasma density
        $n_{0} = 7\nexp{-14}\unit{cm}^{-3}$.}
\end{figure}

>>>>

% ******************************************************************************************************************** %
\section[\label{S:I}]{Introduction}
% interested from SMI/AWAKE
% requirements for AWAKE Run 2
% ******************************************************************************************************************** %

Beam driven plasma wakefield accelerators have the potential to offer compact linear accelerators with high energy
gradients, and has been of interest for several decades \cite{chen:1985}. A relativist beam travelling through a plasma
will excite a strong longitudinal e-field that can be loaded by a trailing witness beam, which can achieve a high
acceleration gradient. Acceleration of an electron beam by an electron drive bunch has been demonstrated experimentally
\cite{rosenzweig:1988, blumenfeld:2007, kallos:2008} in the past. The AWAKE experiment at CERN is a proof of concept
proton driven plasma wakefield accelerator \cite{awake_collaboration:2014}.

A major challenge with plasma wakefield accelerators is, however, to produce an accelerated beam with a minimal increase
in energy spread and emittance. In the well described linear case where the beam density $n_{b}$ is much smaller than
the plasma density $n_{0}$ the plasma causes emittance growth in the beam. A finite length beam will also see a varying
transverse wakefield causing increasing energy spread \cite{katsouleas:1987}. In the non-linear regime, where
$n_{b} > n_{0}$, the drive beam sees an unvarying transverse field due to a bubble forming in the plasma behind the
drive beam. The bubble is formed by the transverse oscillations of the plasma electrons which move near uniformly as a
function of distance from the drive beam forming a sheet around an evacuated area. Since the much heavier plasma ions
have no significant movement within the relevant time frame, these remaining ion channel creates a focusing force that
scales linearly with the radius within this bubble producing a angularly symmetric focusing force
\cite{lu:2006-1, lu:2006}. This effect also avoid the issue of increasing energy spread due to transverse beam loading.
Increase in energy spread is still dependent on beam loading of the longitudinal e-field, $E_{z}$. This effect can be
negated by specially shaped beams \cite{katsouleas:1987, tzoufras:2009}, but this is difficult experimentally and is not
a topic explored in this paper.

% ******************************************************************************************************************** %
\subsection[\label{S:I:SMI}]{Self-modulation as a Driver}
%~ Patric's Comments
%~ - Self-modulation as a driver
%~ - SM does not reach blow out
%~ - SM does not preserve emittance and produces large energy spread and parameters may evolve along the accelerator,
%~   not desirable
% ******************************************************************************************************************** %

A train of electron drive bunches with a separation $\lambda_{pe}$ and a length $l_{b} \ll \lambda_{pe}$, where
$\lambda_{p} = 2\pi c/\omega_{pe}$ and $\omega_{pe} = (n_{0} e^{2} / m_{e} \epsilon_{0})^{1/2}$, will drive an
increasingly strong field $E_{z}$ for each bunch \cite{chen:1985}. A tailing witness beam loading the peak accelerating
phase of this field will quickly gain energy from the wakefield. Acceleration of an electron witness beam driven by two
electron drive beams was demonstrated at Brookhaven National Laboratory \cite{muggli:2011}. However, electron drive
bunches will quickly lose energy to the plasma, and the accelerating witness beam will eventually catch up with the
drive beam and the acceleration stop.

This de-phasing problem can be minimised by instead using a proton drive bunch \cite{caldwell:2009}. Proton beams, like
for instance the LHC and SPS beams at CERN, are ideal candidates for such an accelerator; however, they are orders of
magnitude longer than the plasma wavelengths needed for such applications. This issue is resolved by letting the proton
beam undergo self-modulation before injecting the witness beam into one of the buckets in the modulated structure. This
self-modulation instability is caused by the transverse fields generated by the beam acting upon the beam itself,
causing regions of the beam to rapidly defocus \cite{kumar:2010}. The modulation frequency is close to that of the
plasma, leaving a train of proton bunches along the beam axis.

% ******************************************************************************************************************** %
\subsection[\label{S:I:AWAKE}]{The AWAKE Experiment}
% ******************************************************************************************************************** %

The AWAKE experiment, currently in its first stages of operation at CERN, uses the SPS beam as its driver. The first run
of the experiment is set up with a $10\unit{m}$ rubidium plasma cell, with a nominal plasma density of
$7\nexp{14}\unit{cm}^{-3}$ \cite{gschwendtner:2016}. The corresponding plasma wavelength is $1.26\unit{mm}$. The plasma
density is matched to the SPS proton beam such that $k_{pe}\sigma_{r} = 1$ where $k_{pe} = \lambda_{pe}/2\pi$ is the
plasma wave number. The aim of the first run is to demonstrate self-modulation of the proton beam, and in 2018 to
sample the wake field with a long electron bunch. The study presented here is for run two, which aims to demonstrate
acceleration of a short electron bunch producing at high energy and low emittance and low energy spread.

\begin{figure}[hbt]
    \includegraphics[width=0.99\linewidth,trim={1mm 2mm 1mm 2mm},clip]{figures/figAWAKE}
    \caption{\label{Fig:AWAKER2} A simplified illustration of the experimental setup for AWAKE Run 2. The SPS proton
        beam undergoes self-modulation in the first $4\unit{m}$ plasma cell. The electron witness beam is injected in
        one of the buckets, and undergoes acceleration in the second plasma cell \cite{berglyd_olsen:2015, adli:2016}.}
\end{figure}

The preliminary design of the second run proposes to use two plasma sections, illustrated in figure \ref{Fig:AWAKER2}.
The first section of $4\unit{m}$ is the self-modulation stage. The electron beam will be injected into the modulated
proton beam before stage two, where acceleration will occur. As the $e_z$ field will decrease due to the gap between the
two cells, it is desirable to keep this as short as possible \cite{adli:2016}.

% ******************************************************************************************************************** %
\section[\label{S:M}]{Method}
%- explain analogy of short-bunch, SMI case, by referring to earlier Veronica-papers
%- simulation setup PIC/quickpic OS
% ******************************************************************************************************************** %

The main focus of this study has been on the beam loading of the electron beam. In order to eliminate other factors that
may affect this, we have tried several approaches to create a stable drive beam structure based on previous SMI studies.

Our first approach was to use a pre-modulated, short proton beam with similar structure to the one produced by the
self-modulation of the SPS beam in AWAKE. These studies were done using the full PIC code Osiris \cite{fonseca:2002}
using 2D cylindrical-symmetric simulations. The proton beam was pre-modulated by a clipped cosine function to the
longitudinal density profile, with a period matching $\lambda_{pe}$. Its transverse profile was kept as a Gaussian with
$\sigma_{r} = k_{pe} = 200\unit{\mu m}$. The length of the simulation beam was limited to $26\cdot\lambda_{pe}$, and the
electron beam injected after the 20th micro-bunch \cite{berglyd_olsen:2015}. We performed several parameter scans with
this setup, testing for the the maximum beam charge that could be accelerated with a high energy gain and low energy
spread \cite{adli:2016, berglyd_olsen:2016}.

In order to evaluate the quality of the beam, we also needed to study emittance evolution. However, full PIC codes like
Osiris are vulnerable to numerical growth of emittance caused by the ``numerical Cherenkov effect'' \cite{godfrey:1974}.
This is a know issue with the Yee EMF solver, which causes the phase velocity of electromagnetic fields to be lower than
$c$ while the beam moves very close to $c$. The effect can be mitigated somewhat by replacing the Yee solver with a
solver designed by Lehe \cite{lehe:2013}. These simulations showed that this numerical effect is still prominent in the
high density regions of the electron beam, making it difficult to distinguish the physics from the numerical error.

Quasi-static PIC codes do not suffer this problem, so in order to study the emittance evolution of the beam we instead
turned to the recently released open source version QuickPIC developed by UCLA. QuickPIC a fully relativistic 3D PIC
code \cite{huang:2006, an:2013}.

% ******************************************************************************************************************** %
\subsection[\label{S:M:Setup}]{Drive Beam Parameters}
%~ Patric's Comments
%~ - Self-modulation as a driver
%~ - SM does not reach blow out
%~ - SM does not preserve emittance and produces large energy spread and parameters may evolve along the accelerator,
%~   not desirable
% ******************************************************************************************************************** %

In these simulations we use a single proton drive bunch that sets up a wakefield comparable to that which we expect to
see from the self-modulated SPS beam in AWAKE. The baseline AWAKE drive beam contains $3\nexp{11}$ protons
\cite{gschwendtner:2016}. Only half of these are driving the wakefield, presuming the self-modulation is seeded in the
middle of the unmodulated bunch. In addition. a significant portion of the protons are lost during self-modulation. The
needed gap between the two plasma cells also contributes to a decreased density of protons near the axis, and some of
the field is drained by other protons seeing an accelerating phase within the drive beam. In total, the expected peak
$E_{z}$ field of $~2\funit{GV}{m}$ in the first plasma cell drops to $500-600\funit{MV}{m}$ in the second cell
\cite{awake_collaboration:2016}. A single proton beam of $1.46\nexp{10}$, or $2.34\unit{nC}$ and $7\unit{kA}$, is needed
to drive an $E_{z}$ field of a comparable magnitude.

The single bunch approach was chosen to reduce simulation time for the 3D simulations, and to create a stable
environment for the witness beam. We are not considering the evolution of the proton beam in this test setup, so to
prevent the proton beam from evolving radially, we also increased its mass.

The baseline AWAKE drive beam current is insufficient to reach the non-linear regime and produce a plasma bubble. The
plasma electrons are depleted to around $65\%$ of nominal plasma density at the injection point of the electron beam.
This condition is replicated in out single bunch case as the peak density of the drive beam is $0.83\cdot n_{0}$. The
single bunch setup uses the baseline proton energy $W_{pb} = 400\unit{GeV}$, and retains the transverse size
$\sigma_{r} = k_{pe}$. The length of the bunch $\sigma_{z} = 40\unit{\mu m}$.

% ******************************************************************************************************************** %
\subsection[\label{S:M:Setup}]{Witness Beam Parameters}
% ******************************************************************************************************************** %

The witness beam in our simulation differs from AWAKE baseline parameters on several key points. Initial beam energy is
set such that $\gamma_{eb} = \gamma_{pb} = 426.3$. This was done to eliminate the problem of initial de-phasing of the
witness beam. AWAKE baseline energy for the long witness beam for Run 1 is $W_{eb} = 10-20\unit{MeV}$
\cite{gschwendtner:2016}. Beam loading of a short witness beam is sensitive to initial position. AWAKE Run 2 will likely
require a higher initial energy, but a $W_{eb} = 30-50\unit{MeV}$ is sufficient to minimise de-phasing as the drift
$\Delta\xi \propto \gamma^{-2}$ \cite{berglyd_olsen:2015}. For this case we have eliminated it entirely in order to
reduce the number of variables in our parameter scans.

Earlier simulations with witness beam $\sigma_{r} = 100\unit{\mu m}$ caused the beam radius to undergo rapid transverse
evolution as the beam enters the plasma, reducing its $\sigma_{r}$ from $100\unit{\mu m}$ to $< 10\unit{\mu m}$
\cite{berglyd_olsen:2016}. This is due to a mismatch between beam emittance and the plasma density. This relation is
given by
\begin{align}
    \beta = \frac{\sigma_r^2}{\epsilon_g} = \frac{\lambda_p}{2\pi}\sqrt{2\gamma_{rel}}, \label{EQ:Matched}
\end{align}
where $\lambda_{p}$ is the plasma wavelength. In these simulations we have used matched beams for emittances between
$2$ and $6\unit{\mu m}$, corresponding to $\sigma_{r}$ ranging from $5.25\unit{\mu m}$ to $9.09\unit{\mu m}$.

% ******************************************************************************************************************** %
\section[\label{S:BL}]{Beam Loading}
%~ Eric's Comments
%~ Sec N: Beam loading [in this section we describe the interesting physics, based on a single parameters set]
%~
%~ Patric's Comments
%~ - Beam loading for narrow energy spread
%~ - Beam loading in linear regime (old papers from SLAC in the 80's, then Katsouleas for PWFA) and nonlinear regime
%~   (Tzoufras) well know
%~ - Propose "new regime" in the SM case, beam loading and blowout to get emittance preservation
%~ - Limitation as for PWFA driver, must use some of the W bunch to reach the blowout, it is kind of like the early SLAC
%~   experiments (84GeV) where the W-bunch is like the D-bunch in these experiments, but in a plasma prepared by the p+
%~   bunch and the SM
%~ - maybe also something about the fact that 200pC in a 60µm bunch corresponds to a 1kA current, much larger than the
%~   p+ bunch current, because of the fact that the wakefields are driven my multiple bunches?
% ******************************************************************************************************************** %

\begin{figure}[hbt]
    \includegraphics[width=0.99\linewidth,trim={4mm 0mm 4mm 0mm},clip]{figures/beamLoading}
    \caption{\label{Fig:BeamLoading} Comparison between the unloaded longitudinal e-field (no witness beam) and the
        loaded e-field along the axis. The magnitude of the beam density along the axis is shown for reference. The
        beams travel towards the left. $\xi = z - tc$ is the position in the simulation box. The beam density is in
        units of initial plasma density $n_{0} = 7\nexp{-14}\unit{cm}^{-3}$. The bottom plot shows the plasma density
        for the loaded and unloaded case. The plasma depletion region behind the proton bunch has a minimum of $67.4\%${}
        of $n_{0}$. }
\end{figure}

\begin{figure}[hbt]
    \includegraphics[width=0.99\linewidth,trim={4mm 0mm 4mm 0mm},clip]{figures/plasmaDenTWake}
    \caption{\label{Fig:PlasmaDenTWake} The plasma density in grey with the proton beam (blue) and the electron beam
        (red) superimposed. The line plot indicates the transverse wakefield gradient $dW_{x}/dx$ where
        $W_{x} = E_{x} - v_{b} B_{y}$, evaluated along the beam axis.}
\end{figure}

@@

The single drive beam setup is designed to behave similarly to the self-modulated case. However, since the drive beam is
prevented from significant transverse evolution, we are presented with an idealised case where the electron witness beam
sees consistent fields. The field generated by the proton drive bunch is seen as the dotted lint in figure
\ref{Fig:BeamLoading}. With a proton beam density $n_{pb} \simeq n_{0}$, we are in the quasi-linear regime but near
non-linear \cite{rosenzweig:2010}. There is some depletion of plasma electrons, but not enough to form a bubble. The
dotted green line in the lower part of figure \ref{Fig:BeamLoading} shows that the on-axis plasma density has a
depletion of $67\%$.
